A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-20 tex/example.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

\section{Examples}\label{sec:example}
Our theory works for any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings, and
\item spine calculus, a simply typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first is a standard exercise in implementing a type synthesiser and the second is an example of a system with infinitely many operations.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammars and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
\bgroup
\small
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \\
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

@@

To illustrate this, let us consider the following rule:
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn \textstyle{\prod_{i = 1}^{n}} A_i \bto B \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
The type signature $\Sigma_{\bto}$ can be extended with $\mathsf{fun}_n$ where $\arity({\mathsf{fun}_n}) = n$ for each $n : \N$.
Similarly, the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ can be extended with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
also for each $n : \N$.
Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

>>>>

\section{Examples}\label{sec:example}
Our theory works for any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings, and
\item spine calculus, a simply typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first is a standard exercise in implementing a type synthesiser.
The second example includes not only infinitely many operations but also a type which is not formed by solely one type construct~$\tyOp$.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammars and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we simply adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
\bgroup
\small
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \\
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

@@

To illustrate this, let us consider the following rule:
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
We use the same type signature $\Sigma_{\bto}$ but extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
for each $n : \N$.
Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

### 2023-07-20 tex/sec3-defs.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

\section{Simply Typed Languages and Bidirectional Type Systems}\label{sec:defs}
This section provides generic definitions
%\footnote{%
%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
of simple types, simply typed languages, and bidirectional type systems, using simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
(These definitions may look dense, especially on first reading.
The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.)

The definitions are formulated in two steps:
\begin{enumerate*}
  \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
\item then, given a signature, we define raw terms and typing derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
\end{enumerate*}
All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.

@@

For simple types, the only datum needed for specifying a type construct is its number of arguments:
\begin{definition} \label{def:simple-signature}
  \begin{figure}
      \centering
      \small
      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a type with type variables in $\Xi$ for a signature $\Sigma$}
      \begin{mathpar}
        \inferrule{X_i \in \Xi}{\Xi |-_\Sigma X_i} \and
        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
      \end{mathpar}
      \caption{Simple types}
      \label{fig:simple-type}
  \end{figure}
  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimate sets in the sense of homotopy type theory.
  }
  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
  \begin{enumerate}
    \item a variable in $\Xi$ or
    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
  \end{enumerate}
\end{definition}

\begin{example} \label{ex:type-signature-for-function-type}
  Simply typed $\lambda$-calculus includes function types $A \bto B$ and typically a base type~$\mathtt{b}$ to ensure that the set of all types without type variables is non-empty.
  The type signature $\Sigma_{\bto}$ used to define types in simply typed $\lambda$-calculus consists of a binary operation $\mathsf{fun}$ and a nullary operation $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$. 
\end{example}

>>>>

\section{Simply Typed Languages and Bidirectional Type Systems}\label{sec:defs}
This section provides generic definitions
%\footnote{%
%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
of simple types, simply typed languages, and bidirectional type systems, and uses simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
These definitions may look dense, especially on first reading.
The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.

The definitions are formulated in two steps:
\begin{enumerate*}
  \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory.
  } of operation symbols and an assignment of arities to symbols;
\item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
\end{enumerate*}
%All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.

@@

For simple types, the only datum needed for specifying a type construct is its number of arguments:
\begin{definition} \label{def:simple-signature}
  \begin{figure}
      \centering
      \small
      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a type with type variables in $\Xi$ for a signature $\Sigma$}
      \begin{mathpar}
        \inferrule{X_i \in \Xi}{\Xi |-_\Sigma X_i} \and
        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
      \end{mathpar}
      \caption{Simple types}
      \label{fig:simple-type}
  \end{figure}
  A \emph{signature} $\Sigma$ for simple types consists of a set $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
  \begin{enumerate}
    \item a variable in $\Xi$ or
    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
  \end{enumerate}
\end{definition}

\begin{example} \label{ex:type-signature-for-function-type}
  Function types $A \bto B$ and typically a base type~$\mathtt{b}$ are included in simply typed $\lambda$-calculus and can be specified by the type signature $\Sigma_{\bto}$ consisting of operations $\mathsf{fun}$ and $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$. 
\end{example}

### 2023-07-20 tex/sec4-pre-synthesis.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

\citet[Section~3.2]{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
In our theory, we may formulate annotatability as follows.

>>>>

\citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
In our theory, we may formulate annotatability as follows.
